Combinatorics of the geometry of Wilson loop diagrams II: Grassmann necklaces, dimensions, and denominators

### 2020-12-12 GWLDCI.tex
once through paper 2

There are many other instances of this basic methodology in combinatorics, but we'd like to point out one in particular that involves many objects closely related to those we use here, \cite{CGdyck}.\footnote{Thanks to the referee for the companion paper for pointing this out.}  In particular they, like us, look at a special case within the world of positroids and related objects which is enumerated by Catalan numbers.  For them these are positroids coming from unit interval graphs, while for us they are the equivalent exact subdiagrams.  Our objects are different, which can be seen by the fact that exact subdiagrams do not necessarily satisfy Lemma 4.1 of \cite{CGdyck}, but there are similarities in flavour with our arguments \hlfix{here and in our companion paper.}{Karen still needs to work on this}

>>>>

There are many other instances of this basic methodology in combinatorics, but we'd like to point out one in particular that involves many objects closely related to those we use here, \cite{CGdyck}.\footnote{Thanks to the referee for the companion paper for pointing this out.}  In particular they, like us, look at a special case within the world of positroids and related objects which is enumerated by Catalan numbers.  For them these are positroids coming from unit interval graphs, while for us they are the equivalent exact subdiagrams.  Our objects are different, which can be seen by the fact that the exact subdiagram with 5 vertices (which is unique up to rotation) does not give a decorated permuation in the form of the Main Theorem of \cite{CGdyck}, but there are similarities in flavour with our arguments.

### 2020-12-12 GWLDCII.tex
once through paper 2

In continuing to characterize the positroid cells associated to Wilson loop diagrams, our next goal is to show that the dimension of the positroid cell defined by a Wilson loop diagram $(\cP, [n])$ has dimension $3|\cP|$; this is achieved in Theorem \ref{thm dim}.

@@

Our approach is much more explicit. By combining Algorithms~\ref{alg:put GN on WLD} and \ref{alg:GN to Le} (converting from Wilson loop diagram to Grassmann necklace, and Grassmann necklace to Le diagram respectively) we explicitly describe the effect of adding another propagator to a Wilson loop diagram in terms of the plusses of the associated Le diagrams. Recall that the dimension of a positroid cell is equal to the number of plusses in its associated Le diagram \cite[Theorem 6.5]{Postnikov}; thus understanding the effect of adding another propagator is sufficient to give a recursive proof of the $3|\cP|$-dimensionality of the cells.

@@

\begin{lem}\label{lem good p}
  Let $W$ be an admissible Wilson loop diagram with $k \geq 1$ propagators.  Then there is some dihedral transformation of $W$ such that the resulting diagram $W'$ has a propagator $p$ with the following properties:
  \begin{itemize}
  \item $p = (i, n-1)$ for some $i$, and $p$ has no propagators inside it with regard to the $<_i$ ordering.
  \item Either the edge $i$ in $W'$ only supports $p$, or the edge $i$ in $W'$ supports exactly one other propagator $q = (j,i)$, with no other propagators inside $q$ in the $<_j$ ordering. 
  \end{itemize}
\end{lem}
Note that since the first condition tells us that $p$ has nothing inside it, if $q$ exists in the second condition then $j<_ji<_jn-1$.

@@

\begin{lem}\label{lem I}
Let $W$ be an admissible Wilson loop diagram with $k\geq 1$ propagators, and suppose that $W$ admits one of the configurations described in Lemma~\ref{lem good p}, with $p$ and $q$ (if $q$ exists) both of length 2. Let $V$ be $W$ with $p$ removed.  We can express $I_*^W$ in terms of $I_*^V$ as indicated below:

@@

We note for future reference that the arguments of the previous two paragraphs also imply that $n-2,n-1,n \not\in I_1^V$.

Next consider $I_n^{W}$. From Figure \ref{fig special p} we see that in both cases we have $I_n^{W}(p) = n$; thus from vertex 1 onwards the unassigned propagators are exactly those that appear in $V$. Therefore, the algorithm continues as in $I_1^{V}$, i.e. we have $I_n^{W} = I_1^{V} \cup \{n\}$. By a similar argument, we must have $I_{n-1}^{W} = I_n^{V} \cup \{n-1\}$.

@@

Finally, if $n-2,n-1 \in I_{n-2}^{V}$ then $I_{n-2}^W(q) = n-2$ and $I_{n-2}^W(p) = n-1$ as above, but some other propagator must have contributed $n-1$ to $I_{n-2}^{V}$ so we cannot use the same argument as above. \hlfix{When the algorithm $I_{n-2}^W$}{M.54 wants "When in the algorthm". I disagree.} reaches vertex $n$, only the propagators $p$ and $q$ have contributed; thus we proceed as in the construction of $I_n^{V}$ but without propagator $q$ by Remark \ref{rmk algorithm locally same}. By Lemma~\ref{vertex cyclic int lem} we know that $I_n^V(q) = n-5$, and from Remark \ref{clockwise ordering rem} we see that the only way this could occur is if $q$ was the last to contribute to $I_{n-2}^V$. %all other propagators in $V$ had already been assigned when we reached vertex $n-5$; thus $q$ was the final propagator contributing to $I_n^{V}$. 
Therefore, $I_{n-2}^{W} = (I_{n}^{V} \setminus \{n-5\}) \cup \{n-1, n-2\}$ in this case. This completes all cases for $I_{n-2}^{W}$.

@@

Recall that $I_1$ determines the shape of the Le diagram: specifically, the elements of $I_1$ label the rows of the diagram, working from top right to bottom left. By the arguments in Lemma {\ref{lem I}} we know that ${n,n-1,n-2, n-3 \not\in I_1^{V}}$.  These are the leftmost four columns of the diagram for $V$.  From Lemma \ref{lem I} we know that that $I_1^{W} = I_1^{V}\cup \{n-3\}$. Therefore, the right hand boundary of the shape of $V$ is the same as the right hand boundary of the shape of $W$ except that $W$ has one additional row of 3 boxes while $V$ has an additional column in the $n-3$ position; that is, an extra column fourth from the left. 
\end{proof}
\begin{figure}
  \[\begin{tikzpicture}[scale = 1.5]
\node at (1.5,2.5) {\Large $V$};
\draw (0,0) rectangle (1,2);
\draw (1,0) rectangle (1.3,2);
\begin{scope}[shift = {(-2.7,-2)}]
\draw (4,2) .. controls (5.3,2.2) and (4.7,3) .. (5.5,3.2) .. controls (6.3,3.5) .. (6.5,4) -- (4,4);
\end{scope}
\node at (0.5,1) {\large$\mathcal{A}$};
\node at (1.8,1) {\large$\mathcal{B}$};
\begin{scope}[shift = {(5,0)}]
\node at (1.5,2.5) {\Large $W$};
\draw (0,-0.3) rectangle (1,0);
\draw (0,0) rectangle (1,2);
\begin{scope}[shift = {(-3,-2)}]
\draw (4,2) .. controls (5.3,2.2) and (4.7,3) .. (5.5,3.2) .. controls (6.3,3.5) .. (6.5,4) -- (4,4);
\end{scope}
\node at (0.5,1) {\large$\mathcal{A}$};
\node at (1.5,1) {\large$\mathcal{B}$};
\end{scope}
\end{tikzpicture}\]
  \caption{Le diagrams for $V$ (left) and $W$ (right).}\label{fig Le}
\end{figure}

@@

Specifically with $V$ and $W$ as in Lemma~\ref{lem I} and with $n-2, n-1\in I_{n-2}^V$ (so we are necessarily in Case 2 and $n-1$ supports at least one additional propagator) we have
  \begin{equation}\label{eq necklace}
  I_{n-3}^{V} \xrightarrow[\substack{n-3\text{ out}\\n-2\text{ in}}]{} I_{n-2}^{V} \xrightarrow[\substack{n-2\text{ out}\\n-5\text{ in}}]{} I_{n-1}^{V} \xrightarrow[\substack{n-1\text{ out}\\ u \text{ in}}]{} I_{n}^{V}  \xrightarrow[\substack{n \text{ out}\\v \text{ in}}]{} I_1^{V}
  \end{equation}
where $u$ and $v$ are two vertices; we will derive some properties of them below.

>>>>

In continuing to characterize the positroid cells associated to Wilson loop diagrams, our next goal is to show that the dimension of the positroid cell defined by a Wilson loop diagram $(\cP, [n])$ is $3|\cP|$.  This is achieved in Theorem \ref{thm dim}.

@@

\hlfix{Our approach is}{The transition between these paragraphs has been messed up a bit} much more explicit. By combining Algorithms~\ref{alg:put GN on WLD} and \ref{alg:GN to Le} (converting from Wilson loop diagram to Grassmann necklace, and Grassmann necklace to Le diagram respectively) we explicitly describe the effect of adding another propagator to a Wilson loop diagram in terms of the plusses of the associated Le diagrams. Recall that the dimension of a positroid cell is equal to the number of plusses in its associated Le diagram \cite[Theorem 6.5]{Postnikov}.  Thus understanding the effect of adding another propagator is sufficient to give a recursive proof of the $3|\cP|$-dimensionality of the cells.

@@

\begin{lem}\label{lem good p}
  Let $W$ be an admissible Wilson loop diagram with $k \geq 1$ propagators.  Then there is some dihedral transformation of $W$ such that the resulting diagram $W'$ has a propagator $p$ with the following properties:
  \begin{itemize}
  \item $p = (i, n-1)$ for some $i$, and $p$ has no propagators inside it with respect to the $<_i$ ordering.
  \item Either the edge $i$ in $W'$ only supports $p$, or the edge $i$ in $W'$ supports exactly one other propagator $q = (j,i)$, with no other propagators inside $q$ in the $<_j$ ordering. 
  \end{itemize}
\end{lem}
Note that since the first condition tells us that $p$ has nothing inside it, if $q$ exists in the second condition then $j<_ji<_jn-1$.

@@

\begin{lem}\label{lem I}
Let $W$ be an admissible Wilson loop diagram with $k\geq 1$ propagators, and suppose $W$ admits one of the configurations described in Lemma~\ref{lem good p}, with $p$ and $q$ (if $q$ exists) both of length 2. Let $V$ be $W$ with $p$ removed.  We can express $I_*^W$ in terms of $I_*^V$ as indicated below:

@@

We note for future reference that the arguments of the previous two paragraphs also imply that $n-2$, $n-1$, and $n$ are not in $I_1^V$.

Next consider $I_n^{W}$. From Figure \ref{fig special p} we see that in both cases we have $I_n^{W}(p) = n$. Thus from vertex 1 onwards the unassigned propagators are exactly those that appear in $V$. Therefore, the algorithm continues as in $I_1^{V}$, i.e. we have $I_n^{W} = I_1^{V} \cup \{n\}$. By a similar argument, we must have $I_{n-1}^{W} = I_n^{V} \cup \{n-1\}$.

@@

Finally, if $n-2$ and$ n-1$ are in $I_{n-2}^{V}$ then $I_{n-2}^W(q) = n-2$ and $I_{n-2}^W(p) = n-1$ as above, but some other propagator must have contributed $n-1$ to $I_{n-2}^{V}$ so we cannot use the same argument as above. \hlfix{When, in the algorithm, the construction of $I_{n-2}^W$ reaches vertex $n$,}{M.54 wants "When in the algorthm". I disagree. -- how about this compromise} only the propagators $p$ and $q$ have contributed; thus we proceed as in the construction of $I_n^{V}$ but without propagator $q$ by Remark \ref{rmk algorithm locally same}. By Lemma~\ref{vertex cyclic int lem} we know that $I_n^V(q) = n-5$, and from Remark \ref{clockwise ordering rem} we see that the only way this could occur is if $q$ was the last to contribute to $I_{n-2}^V$. %all other propagators in $V$ had already been assigned when we reached vertex $n-5$; thus $q$ was the final propagator contributing to $I_n^{V}$. 
Therefore, $I_{n-2}^{W} = (I_{n}^{V} \setminus \{n-5\}) \cup \{n-1, n-2\}$ in this case. This completes all cases for $I_{n-2}^{W}$.

@@

Recall that $I_1$ determines the shape of the Le diagram: specifically, the elements of $I_1$ label the rows of the diagram, working from top right to bottom left. By the arguments in Lemma {\ref{lem I}} we know that none of $n$, $n-1$, $n-2$, and $n-3$ are in $I_1^{V}$.  These are the leftmost four columns of the diagram for $V$.  From Lemma \ref{lem I} we know that that $I_1^{W} = I_1^{V}\cup \{n-3\}$. Therefore, the right hand boundary of the shape of $V$ is the same as the right hand boundary of the shape of $W$ except that $W$ has one additional row of 3 boxes while $V$ has an additional column in the $n-3$ position; that is, an extra column fourth from the left. 
\end{proof}
\begin{figure}
  \[\begin{tikzpicture}[scale = 1.5]
\node at (1.5,2.5) {\Large $V$};
\draw (0,0) rectangle (1,2);
\draw (1,0) rectangle (1.3,2);
\begin{scope}[shift = {(-2.7,-2)}]
\draw (4,2) .. controls (5.3,2.2) and (4.7,3) .. (5.5,3.2) .. controls (6.3,3.5) .. (6.5,4) -- (4,4);
\end{scope}
\node at (0.5,1) {\large$\mathcal{A}$};
\node at (1.8,1) {\large$\mathcal{B}$};
\begin{scope}[shift = {(5,0)}]
\node at (1.5,2.5) {\Large $W$};
\draw (0,-0.3) rectangle (1,0);
\draw (0,0) rectangle (1,2);
\begin{scope}[shift = {(-3,-2)}]
\draw (4,2) .. controls (5.3,2.2) and (4.7,3) .. (5.5,3.2) .. controls (6.3,3.5) .. (6.5,4) -- (4,4);
\end{scope}
\node at (0.5,1) {\large$\mathcal{A}$};
\node at (1.5,1) {\large$\mathcal{B}$};
\end{scope}
\end{tikzpicture}\]
  \caption{Le diagrams for $V$ (left) and $W$ (right).}\label{fig Le}
\end{figure}

@@

Specifically with $V$ and $W$ as in Lemma~\ref{lem I} and with $n-2, n-1\in I_{n-2}^V$ (so we are necessarily in Case 2 and $n-1$ supports at least one additional propagator) we have
  \begin{equation}\label{eq necklace}
  I_{n-3}^{V} \xrightarrow[\substack{n-3\text{ out}\\n-2\text{ in}}]{} I_{n-2}^{V} \xrightarrow[\substack{n-2\text{ out}\\n-5\text{ in}}]{} I_{n-1}^{V} \xrightarrow[\substack{n-1\text{ out}\\ u \text{ in}}]{} I_{n}^{V}  \xrightarrow[\substack{n \text{ out}\\v \text{ in}}]{} I_1^{V}
  \end{equation}
where $u$ and $v$ are two vertices. We will derive some properties of $u$ and $v$ below.
